Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model

### 2023-05-22 src/docs_paper/paper.Rmd
491 is the total number of parameters
Update paper writing

## Naomi as an ELGM

@@

@@

We fit the simplified Naomi model (Section \ref{sec:naomi}) to data from Malawi using three inferential approaches.
For each approach, the `TMB` C++ user-template (available in the appendix) used to specify the log-posterior was the same.
The four approaches were:
1. TMB: EB combined with a Gaussian approximation via `TMB`,
2. PCA-AGQH: the PCA AGHQ grid combined with a Gaussian approximation via `aghq`,
3. NUTS: the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using Stan [@carpenter2017stan] via the `tmbstan` package [@monnahan2018no].
The dimension of the latent field is $N = 491$ and the dimension of the hyperparameters is $m = 24$.
Settings used for each inferential method are provided in Table \ref{tab:inference-methods}.
For the deterministic methods, following inference we simulated hyperparameter and latent field samples.
For all methods, we simulated age-sex-district specific HIV prevalence, ART coverage and HIV incidence from the latent field and hyperparameter posteriors.
<!-- Note: eventually this figure should be generated using the algorithm proposed in the paper -->
Example model outputs from `TMB` are illustrated in Figure \ref{fig:naomi-results}.
The \textsc{R} [@r] code used to produce all results we describe below is available at `github.com/athowes/elgm-inf`.
We used `orderly` [@orderly] for reproducible research, `ggplot2` for data visualisation [@wickham2016ggplot2] and `rticles` [@allaire2022rticles] for reporting via `rmarkdown` [@allaire2022rmarkdown].

```{r naomi-results, fig.cap="District-level model outputs for adults aged 15-49. Inference conducted with \\texttt{TMB}."}
knitr::include_graphics("depends/naomi_results.png")
```

@@

We assessed the quality of our NUTS results using the potential scale reduction factor $\hat R$, bulk and tail effective sample sizes, autocorrelation decay plots, univariate traceplots, pairs density plots, and NUTS specific divergent transition and energy assessments.
Full details are provided in the appendix.
We  treat these results from NUTS as a gold-standard to which other inferential methods are compared.

@@

## Inference comparison

@@

We are interested in assessing the accuracy of joint distributions as well as marginal distributions.
Let $\{\bphi_i\}_{i = 1}^n$ be posterior joint samples.
Pareto smoothed importance sampling (PSIS) [@vehtari2015pareto, @yao2018yes].
See a summary of the results in Table and Figure \ref{fig:psis}, and full results available in the appendix.

@@

To write.
See a summary of the results in Table and Figure \ref{fig:mmd}, and full results available in the appendix.

@@

## Case study on exceedance probabilites

@@

The Joint United Nations Programme on HIV/AIDS (UNAIDS) has developed ambitious fast-track targets for scaling up ART treatment with the goal to "end the AIDS epidemic by 2030".
The 90-90-90 by 2020 fast-track target is that 90% of PLHIV know their status, 90% of those are on antiretroviral therapy (ART) and 90% of those have suppressed viral load.
Naomi can be used to assess the probability that the second 90 target has been met, thereby helping to find treatment gaps, by evaluating $\mathbb{P}(\alpha_i > 0.81)$ for each strata $i$.
See a summary of the results in Figure \ref{fig:second90}.

@@

### Finding strata with high incidence strata

Some HIV interventions are cost-effective only within high HIV incidence settings, typically defined as above 1% incidence per year.
Naomi can be used to assess the probability of there being high incidence within a strata by evaluating $\mathbb{P}(\lambda_i > 0.01)$.
See a summary of the results in Figure \ref{fig:1inc}.

@@

We developed an approximate Bayesian inference algorithm motivated by a challenging problem in small-area estimation of HIV in low resource settings.
For the simplified Naomi model in Malawi (Section \ref{sec:results}) our method is demonstrated to be more accurate than TMB, and substantially faster than NUTS.
We anticipate that our method could be added to the Naomi web interface as an alternative to TMB.
Analysts might quickly iterate over model options using the faster, less accurate inference approach, switching to the slower, more accurate approach once they are happy with the results.

@@

In our case study we demonstrated a Bayesian workflow for deterministic inference methods.
We retained the ability to draw samples from the posterior distributions of interest, facilitating use of posterior predictive checks (Section \ref{sec:model-assessment}).

>>>>

## Naomi framed as an ELGM

@@

## TMB

## PCA-AGHQ

@@

We fit the simplified Naomi model (Section \ref{sec:naomi}) to data from Malawi using three inferential approaches.
For each approach, the `TMB` C++ user-template (available in the appendix) used to specify the log-posterior was the same.
The four approaches were:
1. TMB: EB combined with a Gaussian approximation via `TMB`,
2. PCA-AGQH: the PCA-AGHQ grid combined with a Gaussian approximation via `aghq`,
3. NUTS: the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using Stan [@carpenter2017stan] via the `tmbstan` package [@monnahan2018no].
The dimension of the latent field is $N = 467$ and the dimension of the hyperparameters is $m = 24$.
Settings used for each inferential method are provided in Table \ref{tab:inference-methods}.
For the deterministic methods, following inference we simulated hyperparameter and latent field samples.
For all methods, we simulated age-sex-district specific HIV prevalence, ART coverage and HIV incidence from the latent field and hyperparameter posteriors.
<!-- Note: eventually this figure should be generated using the algorithm proposed in the paper -->
Example model outputs from TMB are illustrated in Figure \ref{fig:naomi-results}.
The \textsc{R} [@r] code used to produce all results we describe below is available at `github.com/athowes/elgm-inf`.
We used `orderly` [@orderly] for reproducible research, `ggplot2` for data visualisation [@wickham2016ggplot2] and `rticles` [@allaire2022rticles] for reporting via `rmarkdown` [@allaire2022rmarkdown].

```{r naomi-results, fig.cap="District-level model outputs for adults aged 15-49. Inference conducted with TMB."}
knitr::include_graphics("depends/naomi_results.png")
```

@@

We assessed the quality of our NUTS results to be satisfactory using the potential scale reduction factor $\hat R$ [@gelman1992inference; @vehtari2021rank], bulk and tail effective sample sizes, autocorrelation decay plots, univariate traceplots, pairs density plots, and NUTS specific divergent transition and energy assessments [@betancourt2017conceptual].
Full details are provided in the appendix.
We thus treat the results from NUTS as a gold-standard to which other inferential methods are compared.

@@

## Inference comparison \label{sec:inf-comparison}

@@

As marginal distributions, we are interested in assessing the accuracy of joint distributions.
Let $\{\bphi_i\}_{i = 1}^n$ be samples from a joint posterior.
Pareto-smoothed importance sampling (PSIS) [@vehtari2015pareto, @yao2018yes] is a method for stabilising the ratios $r_s = p(\theta_s, y) / q(\theta)$ used in importance sampling.
Let
$$
p(y \, | \, \mu, \sigma, k) =
\begin{cases}
\frac{1}{\sigma} \left( 1 + k \left( \frac{y - \mu}{\sigma} \right) \right)^{- \frac{1}{k} - 1}, \quad k \neq 0 \\
\frac{1}{\sigma} \exp \left( \frac{y - \mu}{\sigma} \right), \quad k = 0
\end{cases}
$$
be the generalised Pareto distribution with shape parameter $k$ and location-scale parameter $(\mu, \tau)$.
See a summary of the results in Table and Figure \ref{fig:psis}, and full results available in the appendix.

@@

Another way to compare joint distributions is via the maximum mean discrepancy [MMD; @gretton2006kernel].
Let $\Phi = \{\bphi_i\}_{i = 1}^n$ and $\Psi = \{\bpsi_i\}_{i = 1}^n$ be two sets of samples from a joint posterior obtained using different inference methods.
The MMD can be empirically estimated by
$$
\text{MMD}(\Phi, \Psi) = \sqrt{\frac{1}{n^2} \sum_{i, j = 1}^n k(\bphi_i, \bphi_j) - \frac{2}{n^2} \sum_{i, j = 1}^n k(\bphi_i, \bpsi_j) + \frac{1}{n^2} \sum_{i, j = 1}^n k(\bpsi_i, \bpsi_j)}.
$$
See a summary of the results in Table and Figure \ref{fig:mmd}, and full results available in the appendix.

@@

## Case study on exceedance probabilites \label{sec:exceedance}

@@

The Joint United Nations Programme on HIV/AIDS has developed ambitious fast-track targets for scaling up ART treatment with the goal of "ending the AIDS epidemic by 2030".
Specifically, the "90-90-90 by 2020" fast-track target is that:

* 90% of PLHIV know their status,
* 90% of those are on antiretroviral therapy (ART), and
* 90% of those have suppressed viral load.

Naomi can be used to identify treatment gaps by calculating the probability that the second 90 target has been met $\mathbb{P}(\alpha_i > 0.81)$ for each strata $i$.
We found that both TMB and PCA-AGHQ underestimate the probability that the second 90 target has been met in women (Figure \ref{fig:second90}).
This difficulty may be related to interactions between the household survey and ANC components of the model.

@@

### Finding strata with high incidence

Some HIV interventions are cost-effective only within high HIV incidence settings, typically defined as greater than 1% incidence per year.
Naomi can be used to assess the probability of a strata having high incidence by evaluating $\mathbb{P}(\lambda_i > 0.01)$.
We found that both TMB and PCA-AGHQ overestimate these exceedance probabilities (Figure \ref{fig:1inc}).
We do not yet have a working hypothesis as to why this is.

@@

We developed an approximate Bayesian inference algorithm motivated by a challenging problem in small-area estimation of HIV.
For the simplified Naomi model in Malawi (Section \ref{sec:results}) our method is demonstrated to be more accurate than TMB, and substantially faster than NUTS.
We anticipate that our method could be added to the Naomi web interface as an alternative to TMB.
This would enable analysts to quickly iterate over model options using a faster, less accurate inference approach, before switching to a slower, more accurate approach once they are happy with the results.

@@

We demonstrated a Bayesian workflow for deterministic inference methods.
We retained the ability to draw samples from the posterior distributions of interest, facilitating use of posterior predictive checks (Section \ref{sec:model-assessment}) and methods for inference comparison such as Pareto-smoothed importance sampling (Section \ref{sec:inf-comparison}).
The exceedance probabilities case-study (Section \ref{sec:exceedance}) demonstrates the importance of accurate posterior inferences for realistic use-cases of the Naomi model.
